Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-05 AnalyticStructure_main.tex
update

Linear response can be effectively applied to the analysis of the electronic excitations in open systems. In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that 
the ground-state density $\dmnot$ has been provided. The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a 
time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;.
\ee
Here $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
while the \emph{response density operator} $\dm'(\omega)$ describes the modification of the ground-state density induced by the perturbation. In the contest of the LR the response 
density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field.

The purpose of the present analysis is to determine the asymptotic behavior of the main building blocks of the LR in the long range, \emph{i.e.} outside the support domain of the potential. 
%Then we discuss the relation between localization properties and dependence of the results of LR from the specific choice of the basis set and, in particular, from the dimension of the computation domain.

%\subsection{Solution of the Schroedinger equation for open systems}
\subsection{Long range behavior of the eigenvalue solution of the Schroedinger equation}
\label{SEopenSystem}

@@

The argument procedes by splitting the Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term. Since we are dealing with molecular-like systems $\op V$ has a compact support when 
projected in the $\r$-representation. We can formally solve the SE as an inhomogeneous Helmholtz equation by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$. We observe that, due to the properties of $\op V$, the \emph{source term} $\ket{\op V\psi}$ 
of equation \eqref{SEHelmholtz1} is localized in space and, for open boundary conditions in three dimensions, the kernel of $\op G_H(\eps)$ in the coordinate representation reads: 
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\eps) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \eps  < 0\; \\ 
\frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \eps \geq 0
\end{cases} \;,
\ee
where $\alpha = \sqrt{2|\eps|}$. 
This representation of the eigenvalues solution of the SE evidences that the value $\eps=0$ represents the threshold level that determines the asymptotic character of the wave functions. 
Indeed, for $\eps > 0$,  $\op G_H(\eps)$ behaves as a spherical wave and its convolution with the source term  gives raise to a $\psi(\r)$ delocalized in the whole accessible space. Instead, 
for values of $\eps < 0$, the kernel contains an exponential damping factor and the associated wave functions exhibit a bound-state behavior in the long range. Finally, we observe that the 
square-integrability and the continuity of the wave function and its derivatives poses further constrains responsible for the discretization of the energy spectrum for bound-states.

@@

The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition operator, linking the occupied and empty subspaces 
of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm' = \dm_{\perp} \doteq \dmnot\dm'\op Q_0 + \op Q_0\dm'\dmnot
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. Due to this fact it is possible to introduce an explicit representation of the 
response density as follows 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Here we have defined the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, that span the empty subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equations \cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm']$ due to the density-dependence of $\hnot$.

@@

The properties of the Liouvillian superoperator (see appendix?) imply that excitations are transverse 
operators so that $\op E_a = \dmnot\op E_a\op Q_0 + \op Q_0\op E_a\dmnot$. This condition allows us to introduce a dyadic representation of excitations as follows
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;,
\end{align}
where the \emph{excited states} $\ket{\phi^E_p}$, $\bra{\chi^E_p}$ span the empty subspace of $\hnot$.

@@

Susceptibility written as
\be\lb{RhopExcitationDef2}
\sop{{\cal \chi}}(\omega) \cdot   = \left(\sum_a + \int\dd a   \right) \op W_{a} \,
\frac{2\trace{\op W_a \cdot}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\op W_a = \frac{1}{\sqrt{2}}(\op E_a + \op E_a^t)$. 
% The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
% Liouvillian. Pursuing this procedure provides:
% \be\lb{RhopExcitationDef1}
% \sop{{\cal R}}(\omega) \cdot   =
% \left(\sum_E + \int\dd E   \right) \op E \,
% \frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;,
% \ee 
% where we have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef1}. 
% The elements of the first sector 
% correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

@@

From the theoretical analysis of the previous section we can made two statements:
\begin{itemize}
 \item Fluctuation states, which constitute the building blocks of the response density, have a bound state-like character up to $\omega=|\eps_h|$ and behave as unbound delocalized
 wave functions for higher values of $\omega$. 
 \item The excitation landscape is composed of two sectors with distinct properties. The elements of the first one behave as localized, observable quantities and contribute as discrete 
 poles to the spectral decomposition of the susceptibility functional. The second sector contains a continuum of delocalized excitations which represents a tool to provide
 an effective representation of the spectral function. 
\end{itemize}
The topic to be discussed in this section is the following: on the basis of these statements can we define some indicator to evaluate the reliability of a linear response calculations? And, 
if so, up to which energy threshold can we assume that our computations do converge?

@@

@@

%\subsection{Results for the benzene molecule}

>>>>

Linear response can be effectively applied to the analysis of the electronic excitations in open systems. In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that 
the ground-state density $\dmnot$ has been provided. The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a 
time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
and we have introduced the \emph{response density operator} $\dm'(\omega)$ that codifies the modification of the ground-state density induced by the perturbation.

The main objective of this section is to investigate the structure of the building blocks needed to express the response density operator. In particular, we are interested in clarifying 
the behavior of the main actors of the LR in the long range, \emph{i.e.} outside from the support domain of the potential. 
%Due to this analysis we aim to be able to understand if the objects we are dealing with behave as bound-state like functions or if 
%This analysis allows us to establish comprehension of the type of objects one is dealing with when performing a LR calculation and

%in determining the long range behavior, \emph{i.e.} outside the support domain of the potential. This analysis  
%Then we discuss the relation between localization properties and dependence of the results of LR from the specific choice of the basis set and, in particular, from the dimension of the computation domain. 
%asymptotic behavior of the main building blocks of the LR in the

\subsection{Long range behavior of the eigenvalue solution of the Schroedinger equation}
\label{SEopenSystem}

@@

The argument proceeds by splitting the Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term. Since we are dealing with molecular-like systems $\op V$ has a compact support when 
projected in the $\r$-representation. We can formally solve the SE as an inhomogeneous Helmholtz equation by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$. We observe that, due to the properties of $\op V$, the \emph{source term} $\ket{\op V\psi}$ 
of equation \eqref{SEHelmholtz1} is localized in space and, for open boundary conditions in three dimensions, the kernel of $\op G_H(\eps)$ in the coordinate representation reads: 
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\eps) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \eps  < 0\; \\ 
\frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \eps \geq 0
\end{cases} \;,
\ee
where $\alpha = \sqrt{2|\eps|}$. 
This representation of the eigenvalues solution of the SE evidences that the value $\eps=0$ represents the threshold level that determines the asymptotic character of the wave functions. 
Indeed, for $\eps > 0$,  $\op G_H(\eps)$ behaves as a spherical wave and its convolution with the source term  gives raise to a $\psi(\r)$ delocalized in the whole accessible space. Instead, 
for values of $\eps < 0$, the kernel contains an exponential damping factor and the associated wave functions exhibit a bound-state behavior in the long range. Finally, we observe that the 
square-integrability and the continuity of the wave function and its derivatives poses further constrains responsible for the discretization of the energy spectrum for bound-states.

@@

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm' = \dm_{\perp} \doteq \dmnot\dm'\op Q_0 + \op Q_0\dm'\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. Due to this fact it is possible to introduce an explicit representation of the 
response density as follows 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Here we have defined the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, that span the empty subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equations \cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm']$ due to the density-dependence of $\hnot$.

@@

The properties of the Liouvillian superoperator (see appendix?) imply that excitations are transverse 
operators so that $\op E_a = \dmnot\op E_a\op Q_0 + \op Q_0\op E_a\dmnot$. This condition allows us to introduce a dyadic representation of excitations as follows
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;,
\end{align}
where we have defined the \emph{excited states} $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ that, by construction, are orthogonal to the occupied orbitals of $\hnot$.

@@

Susceptibility written as
\be\lb{RhopExcitationDef2}
\sop{{\cal \chi}}(\omega) \cdot   = \left(\sum_a + \int\dd a   \right) \op W_{a} \,
\frac{2\trace{\op W_a \cdot}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\op W_a = \frac{1}{\sqrt{2}}(\op E_a + \op E_a^t)$. 
We have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef2}. 
The elements of the first sector 
correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

\subsubsection*{Results of this section}

From the theoretical analysis of the previous section we can made two statements:
\begin{itemize}
 \item Fluctuation states, which constitute the building blocks of the response density, have a bound state-like character up to $\omega=|\eps_h|$ and behave as unbound delocalized
 wave functions for higher values of $\omega$. 
 \item The excitation landscape is composed of two sectors with distinct properties. The elements of the first one behave as localized, observable quantities and contribute as discrete 
 poles to the spectral decomposition of the susceptibility functional. The second sector contains a continuum of delocalized excitations which represents a tool to provide
 an effective representation of the spectral function. 
\end{itemize}

% The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
% Liouvillian. Pursuing this procedure provides:
% \be\lb{RhopExcitationDef1}
% \sop{{\cal R}}(\omega) \cdot   =
% \left(\sum_E + \int\dd E   \right) \op E \,
% \frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;,
% \ee 
% where we have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef1}. 
% The elements of the first sector 
% correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

@@

% From the theoretical analysis of the previous section we can made two statements:
% \begin{itemize}
%  \item Fluctuation states, which constitute the building blocks of the response density, have a bound state-like character up to $\omega=|\eps_h|$ and behave as unbound delocalized
%  wave functions for higher values of $\omega$. 
%  \item The excitation landscape is composed of two sectors with distinct properties. The elements of the first one behave as localized, observable quantities and contribute as discrete 
%  poles to the spectral decomposition of the susceptibility functional. The second sector contains a continuum of delocalized excitations which represents a tool to provide
%  an effective representation of the spectral function. 
% \end{itemize}
The topic to be discussed in this section is the following: on the basis of these statements can we define some indicator to evaluate the reliability of a linear response calculations? And, 
if so, up to which energy threshold can we assume that our computations do converge?

@@

\subsection{Fluctuation state and excitations}

Express the fluctuation state in terms of excitation using the relation $\ket{f_p(\omega)} = \dm'(\omega)\ket{\psi_p}$ and using formula \eqref{SusceptibilityDef1} to express the excitation
expansion of the response density. Due to this procedure we obtain
\be\lb{FSExcitationDef1}
\ket{f_p(\omega)}   = \left(\sum_a + \int\dd a   \right) \ket{w_p^a} \,
\frac{\bra{\psi_p}\op\Phi(\omega)\ket{w_p^a}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$. 
Show the convergence of the statical polarizability of the $CO$ in function of the number of excitations. Write formula
\be\lb{AlphaExcitationDef1}
\alpha_{xx} =  \left(\sum_a + \int\dd a   \right) 2 \,
\frac{\left[\trace{\op W_a \op X}\right]^2}{\Omega_a^2} \;,
\ee 
and add the plot that show the convergence rate of the statical polarizability to the reference (finite difference) value in functions of the number of excitations. The plot evidence that the
continuum sector of the excitation spectrum is 
\begin{figure}
\includegraphics[scale=0.56]{CO_alphaX_isto.eps}
\caption{\label{co_AlphaExc}(Color online) Convergence of the statical polarizability of the $CO$ molecule w.r.t the number of excitations.}
\end{figure}

@@
